Quantifying unmet ICU beds in tertiary hospitals in urban India using Propensity

### 2022-10-01 manuscript.Rmd
efter johannas kommentarer

---
title: 'Quantifying unmet ICU beds in tertiary hospitals in urban India using Propensity
  Score Matching '
author: "Maria Eriksson"
output:
  pdf_document: default
  html_document: default
  word_document: default
bibliography: bibliography.bib
csl: vancouver.csl
---

@@

@@

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->
The eligibility criteria for TITCO were patients who presented with injuries from trauma at the emergency departments of the participating hospitals with a mechanism of road traffic, railway, fall, burn or assault. Patients who were dead on arrival were excluded, as were patients who only had isolated limb injuries.

@@

<!-- Describe any efforts to address potential sources of bias -->
The TITCO dataset are observational data collected from a clinical environment and not an experimental setting.  Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patient's medical condition. There exists a causal effect between the independent and the dependent variables: a patient is sent to the ICU precisely because of the severity of their condition.

There exists a potential bias in the routines, resources and patient makeup of the participating hospitals. Because they are all tertiary hospitals, about two thirds of the patients have been referred from other health care providers, meaning that there is a selection bias in the patient populations. Further, hospitals with dedicated trauma wards have a propensity to receive transferred or police escorted patients with worse condition and prognosis.

As stated in chapter Settings, most patients have low socioeconomic status. Low socioeconomic status is associated with higher levels of malnutrition and comorbidities and would have a negative effect on vital parameters and the capacity of patients to tolerate both injuries and interventions. This circumstance doesn't affect the analysis and internal validity, but reduces the generalizability of the results.

Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but did not, these particular selection biases are not expected to skew the outcome, and therefore does not have to be addressed.

Hospitals with trauma wards could have lower admission criteria to the ICU than hospitals with no trauma ward. Because examining admission criteria to the ICU is part of the analysis, any bias from differences in hospital admission criteria must be minimized. This is why the PSM model is trained using a training and validation subset as described in the chapter Statistical methods. If this proves insufficient and unsatisfactory it may be necessary to examine and address this bias by comparing ICU admitted patients between hospitals using the same PSM methodology, but this will go beyond the time frame and scope of this thesis.

@@

Study size 
----------

<!-- Explain how the study size was arrived at -->
Data of 16000 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings.

@@

(a) Describe all statistical methods, including those used to control
for confounding
-->
Propensity score matching (PSM) is useful method for identifying and assembling a control group from within a data set where there is causality between the independent and dependent variables, and a randomized case-control (RCT) study to offset or balance this bias is impractical, immoral or otherwise improper. Propensity score is a statistical variable, understood as the likelihood of a subject receiving treatment conditional on observations of relevant covariates of that subject before treatment. In RCTs it is known from study design, but it is generally unknown in observational data. The rationale for PSM is that while it is unknown, this variable can be fairly well estimated using logistic regression.

The theory being examined in this master's thesis is that there exist a subset of the TITCO dataset that did not go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention. The thesis is that these patients were not admitted to the ICU only for lack of such resources, not by a medical assessment. 
Translated into this context this means that in this data set exist a selection bias, there is causality between the independent variables medical conditions and interventions and the dependent variable ICU admittance. Patients were admitted to the ICU as a consequence of their medical condition. Performing a randomized case-control study and randomly assigning or withholding a live-saving treatment would obviously be immoral, but we believe that in the TITCO data set exist patients who were not admitted into the ICU not by volition but for scarcity of such resources. Such a group would be akin to an non-experimental control group and PSM can be used to identify and assemble it.  If such a subset can be identified by PSM, that is indicative there exists shortage of ICU beds in tertiary hospitals in India.

The proportion of the subset of patients that should hade received ICU care based on our prediction model versus the subset that did receive ICU care is also the proportion of ICU bed shortage. If, for every patient that did receive ICU care, there was another patient with similar predictors that should hade received ICU care based on our prediction model but did not due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.

The propensity score model is developed using the Training Data and tested using the Validation Data. The workflow contains a series of steps. The output of each step is the input for later steps. The model is developed using Testing Data and is tested with Validation Data. By reverting back and adjusting assumptions in an iterative process, the model is continuously improved and the accuracy and validity is improved.

The first step of PSM is to randomly split the data set into two parts, 60% and 40%. The 60% part is Training Data used to develop the propensity score model, and the 40% part is the Validation Part. This is done because the inherent bias of the PSM method. The prediction model produced in step 3 is based on and fitted against the Training data and predictors are weighed so that the the model awards patients that did go to the ICU corresponding high propensity. This model can then be used to predict new patients propensity to go to the ICU based on their baseline predictor values, and by saving 40% in the Validation data set we have a set of patients that is new to the model on which we may test how well the model predicted their fate. 
Training Data is used in steps 2-3 and the resulting model is then validated using Validation Data.

@@

Third step is to build a multivariable logistic regression model with ICU care as the outcome and the predictors selected in the second step as independent variables using Testing Data.

Fourth step is to apply this model on patients from the Validation Data set, and see whether the model correctly prediced which patients in the Validation data received ICU treatment. Using the model, the standard variable propensity score is calculated for each patient in the Validation Data set. If the patients that were admitted to the ICU in the validation set receive appropriate propensity scores by the model constructed in step 2-3, the model is valid. The propensity score determines the likelihood that a patient will receive ICU care, given their condition. 
<!-- 
#fråga: är det såhär man gör menade du Johanna? vad är det egentligen som händer i steg 3 och vad består "träningen" av i steg 2-3? År det att man har valt sina variabler i steg 2 men vi vet inte om de är bra och vi vet inte deras inbördes relevans/vikt för den oberoende variabeln. Så vi petar in dem i steg 3 och klickar OK, och sen bygger R en multivariabelregressionsmodell åt en som viktar prediktorerna åt en så att modellen tilldelar patienterna propensity scores som stämmer bäst överens med deras ICU-status (som är känt för R). R tweakar/vikar/manipulerar modellen att passa träningsdatasetet man matat den med, och bygger en modell som mha prediktorerna är bäst på att identifiera de patienter som R vet hamnade på ICU?
Det gör vi för att optimera modellens kapacitet att "förutse" vad som kommer hända med en "helt ny" patient med en uppsättning värden på prediktorerna? 
Så har vi i Validation Data sparat en uppsättning "helt nya" patienter som vi prövar modellen på? Och ser om modellen är bra på att ge ICU-patienter rimliga propensity scores?
Är det det du menar att modellen ska "träna" på ett set och testas på ett annat?
Och om man switchar över redan i steg 4 finns det ingen validering sen mera för dålig PSM i steg 5? -->

Fifth step is PSM. In PSM members of the subset that did not get ICU care but had similar predictors - and therefore similar propensity scores - as the subgroup that did get ICU care, are matched to patients who did go to the ICU. There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable. There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria and inadvertently vastly under- or overestimate the  need of ICU beds remain.  
<!-- (b) Fråga: 1. visst finns dessa svagheter i propensity score matching?
2. vad gör man åt dem? 
3. är beskrivningen av åtgärder mot detta något som hör hemma här eller i results?-->

The proportion of the subset of patients that probably should hade received ICU care based on our prediction model versus the subset that did receive ICU care is also the proportion of ICU bed shortage.

>>>>

---
title: 'Quantifying unmet ICU beds in tertiary hospitals in urban India using Propensity
  Score Matching '
author: "Maria Eriksson"
output:
  word_document: default
  html_document: default
  pdf_document: default
bibliography: bibliography.bib
csl: vancouver.csl
---

@@

<!-- 
Definitions
------------
An ICU bed is a tiltable hospital bed but included in the term is also staffing and auxiliary equipment. The equipment is to administer oxygen and to monitor vitals. Some ICU beds are equipped with ventilators and if applicable length of ventilation has been registered.

Staffing is one doctor in the ward, and 1:4 nurses per patient but there is not always strict adherence to staffing guidelines. 
If the ICU ward is 100% full, patients might receive ventilation outside the ICU ward. Though the care these patients receive could very well be considered ICU care, they are not registered as such in the TITCO registry.
-->

@@

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->
The eligibility criteria for TITCO were all patients who presented with injuries from trauma at the emergency departments of the participating hospitals with a mechanism of road traffic, railway, fall, burn or assault. Patients who were dead on arrival were excluded, as were patients who only had isolated limb injuries.

@@

<!-- Describe any efforts to address potential sources of bias -->
The TITCO dataset is observational data collected from a clinical environment and not an experimental setting.  Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patient's medical condition. There exists a causal effect between the independent and the dependent variables: a patient is sent to the ICU precisely because of the severity of their condition. There is also a selection bias in which parameters are recorded and which examinations are performed, introducing detection and observer bias.

There exist a bias in the routines, resources and patient makeup of the participating hospitals. Because they are all tertiary hospitals, about two-thirds of the patients have been referred from other health care providers, meaning that there is a selection bias in the patient populations. Further, hospitals with dedicated trauma wards have a propensity to receive more transferred or police-escorted patients.

As stated in chapter Settings, most patients have low socioeconomic status. This circumstance doesn't affect the analysis and internal validity, but reduces the generalizability of the results.

Biases like these are expected in data collected in a clinical setting, the analysis model i chosen to adress this. Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but did not, these particular selection biases are not expected to skew the outcome, and therefore does not have to be addressed.

Hospitals with trauma wards could have different admission criteria to the ICU than hospitals with no trauma ward. The same goes for all equipment any hospital has as well as the proficiency of their physicians. Because examining admission criteria to the ICU is part of the analysis, any bias from differences in hospital admission criteria must be minimized. This is why the PSM model is trained using a training and validation subset as described in the chapter Statistical methods. If this proves insufficient and unsatisfactory it may be necessary to examine and address this bias by comparing ICU admitted patients between hospitals using the same PSM methodology, but this will go beyond the time frame and scope of this thesis.

@@

@@

(a) Describe all statistical methods, including those used to control
for confounding
-->
Propensity score matching (PSM) is useful method for identifying and assembling a control group from within a data set where there is causality between the independent and dependent variables, and a randomized case-control (RCT) study to offset or balance this bias is impractical, immoral or otherwise improper.

The theory being examined in this master's thesis is that there exist a subset of the TITCO dataset that did not go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to be admitted. The thesis is that these patients were not admitted to the ICU only for lack of ICU resources, not by a medical assessment. Such a group would be akin to an non-experimental control group that PSM can be used to identify and assemble.

If such a group of patients can be identified by PSM, that is indicative there exists shortage of ICU beds in tertiary hospitals in India. The size of this group versus the size of the group that did receive ICU care indicate the magnitude of the ICU bed shortage. If, for every patient that did receive ICU care, there was another patient with similar predictors that should hade received ICU care based on our prediction model but did not due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.

The propensity score model is developed using the Training Data and tested using the Test Data. The workflow contains a series of steps. The output of each step is the input for later steps. Steps three and four that involve development and validation of a multivariable regression model have several standard methods to test and improve model validity, and the matching process of step five have several means by which you can control and validate the process. By reverting back and adjusting assumptions in an iterative process, the model is continuously improved and the accuracy and validity is improved.

The first step of PSM is to randomly split the data set into two parts, 60% and 40%. The 60% part is Training Data used to develop the propensity score model, and the 40% part is the Test Part. This is done to detect overfitting of the model to the Training Data.
Training Data is used in steps 2-3 and the resulting model is then validated using Test Data.

@@

Third step is to build a multivariable logistic regression model with ICU care as the outcome and the predictors selected in the second step as independent variables. The prediction model is based on the Training Data and predictors are weighed so that the the model awards patients that did go to the ICU corresponding high propensity. This model can then be used to predict new patients propensity to go to the ICU based on their baseline predictor values. By saving 40% in the Validation data set we have a set of patients that is new to the model. This allows to test how well the model correctly predicted ICU patients by awarding them high propensity scores.

Fourth step is to apply this model on patients from the Test Data. Using the model, the standard variable propensity score is calculated for each patient in the Test Data set. The propensity score determines the likelihood that a patient will receive ICU care, given their condition. If the model is suitable, it will correctly identify patients that did go to the ICU by awarding them high propensity scores, but non-ICU patients that were similar to this cohort will also get high propensity scores.

<!-- 
fråga: vi är ju ute efter ge vissa non-ICU pat höga propensity scores för att sedan kunna identifiera och filtrera fram dem i propensity score matchingen steg 5. Men hur identifierar man skillnaden mellan en patient som förmodligen borde fått ICU, och en patient modellen felaktigt spottat ut?
Och om man switchar över redan i steg 4 finns det ingen validering sen mera för dålig PSM i steg 5? -->

Fifth step is PSM. In PSM members of the subset that did not get ICU care are matched to patients who did go to the ICU, matching based on their propensity scores. There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable. There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria and inadvertently vastly under- or overestimate the  need of ICU beds remain.  
<!-- (b) Fråga: 1. visst finns dessa svagheter i propensity score matching?
2. vad gör man åt dem? 
3. är beskrivningen av åtgärder mot detta något som hör hemma här eller i results?-->
